A sustainable approach to tourist signage on (historical) trails

### 2023-06-29 paper.tex
Fatta prima versione bibliografia

%=================================================================
\documentclass[journal,article,submit,pdftex,moreauthors]{Definitions/mdpi}

@@

Although tourism persistently targets most renowned sites, like art cities and parks, there is an increasing interest for targets not exposed to mass tourism, that may provide a more intense and inner experience. Suh an exclusive flavor is appreciated not only by the occasional visitor, which helps the economy of the corner, but also by those that have their homes in such privileged location. The key to sustain such a difficult equilibrium is to guide the visitor to enjoy the place with discrete messages that he finds on his way.

%The introduction should briefly place the study in a broad context and highlight why it is important. It should define the purpose of the work and its significance. The current state of the research field should be reviewed carefully and key publications cited. Please highlight controversial and diverging hypotheses when necessary. Finally, briefly mention the main aim of the work and highlight the principal conclusions. As far as possible, please keep the introduction comprehensible to scientists outside your particular field of research. Citing a journal paper \cite{ref-journal}. Now citing a book reference \cite{ref-book1,ref-book2} or other reference types \cite{ref-unpublish,ref-communication,ref-proceeding}. Please use the command \citep{ref-thesis,ref-url} for the following MDPI journals, which use author--date citation: Administrative Sciences, Arts, Econometrics, Economies, Genealogy, Humanities, IJFS, Journal of Intelligence, Journalism and Media, JRFM, Languages, Laws, Religions, Risks, Social Sciences, Literature.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods}

@@

%=====================================
% References, variant A: external bibliography
%=====================================
%\bibliography{your_external_BibTeX_file}

%=====================================
% References, variant B: internal bibliography
%=====================================
\begin{thebibliography}{999}
% Reference 1
\bibitem[Author1(year)]{ref-journal}
Author~1, T. The title of the cited article. {\em Journal Abbreviation} {\bf 2008}, {\em 10}, 142--149.
% Reference 2
\bibitem[Author2(year)]{ref-book1}
Author~2, L. The title of the cited contribution. In {\em The Book Title}; Editor 1, F., Editor 2, A., Eds.; Publishing House: City, Country, 2007; pp. 32--58.
% Reference 3
\bibitem[Author3(year)]{ref-book2}
Author 1, A.; Author 2, B. \textit{Book Title}, 3rd ed.; Publisher: Publisher Location, Country, 2008; pp. 154--196.
% Reference 4
\bibitem[Author4(year)]{ref-unpublish}
Author 1, A.B.; Author 2, C. Title of Unpublished Work. \textit{Abbreviated Journal Name} year, \textit{phrase indicating stage of publication (submitted; accepted; in press)}.
% Reference 5
\bibitem[Author5(year)]{ref-communication}
Author 1, A.B. (University, City, State, Country); Author 2, C. (Institute, City, State, Country). Personal communication, 2012.
% Reference 6
\bibitem[Author6(year)]{ref-proceeding}
Author 1, A.B.; Author 2, C.D.; Author 3, E.F. Title of presentation. In Proceedings of the Name of the Conference, Location of Conference, Country, Date of Conference (Day Month Year); Abstract Number (optional), Pagination (optional).
% Reference 7
\bibitem[Author7(year)]{ref-thesis}
Author 1, A.B. Title of Thesis. Level of Thesis, Degree-Granting University, Location of University, Date of Completion.
% Reference 8
\bibitem[Author8(year)]{ref-url}
Title of Site. Available online: URL (accessed on Day Month Year).
\end{thebibliography}

>>>>

%=================================================================
\documentclass[sustainability,article,submit,pdftex,moreauthors]{Definitions/mdpi}

@@

Although an official decision is still pending, there is a sound proposal of naming the time we are living as "anthropocene". Behind such a proposal is the recognition that humans have the potential to significantly alter the environment they live in at a global scale.

The problem behind the above fact is that we humans are not able to understand the consequences of the deployment of such a potential, and not even to keep it under control. Even worse, such capabilities are often used aiming at short-term local effects, ignoring long-term global ones, with the risk of a long-term global deterioration. Such an attitude pervades all activities, starting from the most basic ones like the provision of food and homes, and is frequenly biased by economical profit.

Leisure activities play a relevant role, since they bring profit to niche businesses. This creates a reinforcement effect which follows the negative dynamic seen above: a leisure activity that per se is neutral (e.g. skying) may have a relevant footprint due to induced effects (ski lifts) which reinforces the success of the activity (easier to practice) bringing more investments and impact. Unlike beneficial effects, which is usually well localized, the negative impact has a very wide spectrum encompassing environmental, social and finally economical domains, as discussed by Patthey in \cite{pat08a}.

In an effort to manage such a situation, international agencies promote the "sustainability" keyword as a guideline to the conservation of our habitat. Jeffry Ramsey et al. \cite{ram15a} consider the concept behind the word is "vague and contested but not meaningless" and that it cannot be used without a concrete framework, especially in a normative context. To solve to problem of measuring sustainability, Tom Kuhlman et al. point out a inherent tension in the term between growth and stability. A tension that has been softed by a shift in meaning of the word, which moved from focusing on preservation for an indefinite future to a concept encompassing three "pillars", society, economy and environment, as determinants that need to be protected in the present in view of a future sustainable development. The authors put into relationship man-made and environmental assets, discriminating the two cases in which a loss in the latter may or may not be compensated by an increment of the former, introducinging the terms of {\em weak} and {\em strong} sustainability.

The promotion of the human cultural heritage has its stand in such a complex scenario. As noted by Jorge Otero in \cite{ote22a}, local communities take advantage of heritage promotion: the preservation activity generates social growth, while tourism provides an extra income that may be re-invested in promotion. However, just like in the sky resorts case mentioned above, the risk of compromising the resource itself exists. Such an interplay rises a question of sustainability even for cultural preservation initiatives, two topics that are usually regarded as in synergy. As in the "World Heritage and Sustainable Development" document by Unesco \cite{lab17a} that calls for actions that "harness the potential of World Heritage properties and heritage in general, to contribute to sustainable development".

This paper originates from such a painful contradiction trying to give an answer to the question: is there a way to unlock the potential of a cultural resource limiting the damage to a social, economic and environmental equilibrium?

The question arises in the course of a project funded by the italian ministry of reasearch aiming at the application of cutting edge technologies to the investigation of karst caves that may have been inhabited in the past, from prehistory to today. Among the results expected from the project there is the definition of a strategy to disseminate research results, and allow non-specialists to visit and enjoy the findings, thus contributing to the economy of rural areas that are severely exposed to de-population. A sustainable strategy is even more crucial given that the geological enviroment of interest is notably fragile, as noted by Aleksander Antic\`c et al. in \cite{ant20a}.

This paper provides a concept solution of a specific aspect. The solution is conceptual in that it tries to follow a guideline that takes into account as many side details as possible in a way that can be used in other milieu, but very practical in its unfolding. Given the tighly interconnected nature of the situation, we observe how a vary specific aspect, the touristic signage, exhibits effects on the society, economicy and environment.

\section{Unlocking site potentials through signage}

The signage used to guide and inform the visitor is of paramount importance and must be carefully designed in order to accomplish its functions while following the guidelines of a sustainable approach.

Its functions are twofold
\begin {itemize}
\item an effective signage must guide the tourist across the resort. A map represents a starting point to this end, but not all visitors feel comfortable reading a map. The ideal signage should include visible reference points and explicit actions, like {\em turn right after crossing the stream}. A good example of such a signage is the one used by the Italian Alpine Club (CAI): it consist of a colored stripes marked on tree bark. The marks are placed in such a way that, standing nearby one of them, the hiker is able to see the successive on the trail
\item a sign should explain the reasons of interest of a site. This includes featuring the site where the sign is installed, but also other nearby points of interest that invite the visitor to continue the visit. In this sense, the CAI signal above is not of use. A simple board with a site name may be sufficient only when the site is sufficiently reknown. Otherwise, a more structured message should be used to explain the reasons why the feature is regarded as relevant.
\end{itemize}

In our study we consider a range of solutions that help towards a sustainable support to a sustainable tourism, and specifically to the dynamic provision of information during a visit along a natural trail, not covered by Internet connectivity (a {\em dead zone} using the term used by Pearce and Gretzel \cite{pea12a}). We consider the presence of two stakeholders for the signage itaself \cite{wan22a}: the hosting organization (the {\em host} for short) which implements a signing installation, and the visitor (or {\em user} ) which extracts useful and enjoyable information from signing devices.

The purpose of such an installation is guiding the {\em user} in a tour that includes urban streets, buildings, natural trails and caves (this latter being the topic of the project giving financial support to our research). The task of the {\em host} organization is to provide the {\em user} with all sorts of information that may guide him across the visit, making it as much profitable and enjoyable as possible.

\begin{itemize}
	\item dimensions: the board must be sufficiently large to contain the desired content, taking into account its readability from a distance
	\item installation logistics: depending on the location, the transportation and placement of a plaque may require a basement or other sorts of supports
	\item environmental impact: in order to be effective, the plaque has to be prominent, and this may negatively affect the quality of the site 
	\item accessibility for the visually impaired: the plaque is not useful for visually impaired persons
	\item accessibility for stranger visitors: to limit the size of the plaque, the number of translations must be limited as well
	\item update limits: to update the content, the plaque must be replaced
	\item removal logistics: when the board degrades it must be either removed or replaced, which entails waste disposal together with other issues similar to those found during the installation
\end{itemize}

\subsection{Technology to minimize intrusion}

We characterize the problem as an instance of {\em weak sustainability}: we do not preclude intervention on the habitat, but the impact of such intervention must be better than that related with traditional signs. To this end we include in our solution tools that are not part of the site but remain with the user.

During the last decades, we witnessed the diffusion of smartphone devices, and we have no reasons to expect a change in such a trend. Smartphones empower individual communication capabilities, allowing to receive sound, visual and tactile interactions. The relevance of such capabilities for the improvement of a tourist experience have been widely investigated, etiher in urban environments \cite{liu16a} or in a rural milieu \cite{kum20a}. to investigate ways to exploit such tools in our context. Also in this case we start from the limits of such technology, especially in the case of outdoor activities.

One is related with the diffusion of such a technology, which, although widely available, is not equally familiar to everybody. This is related with {\em usability}, with a term borrowed from P. Wan in \cite{wan22a}. The second is that several functions depends on the provision of enabling services: for instance, their networking capability is useless if the device cannot reach an enabling infrastructure. So the {\em applicability limits} of a solution need to be defined.

Starting from the two aspects above we envision the directions towards a successful smartphone-based strategy, keeping in mind the overall sustainability of the approach.

Regarding {\em usability}, the basic recommendation is to keep the operation as simple as possible, within the experience of the majority of users, without requiring the need of installing new applications.

The definition of its {\em applicability limits} is more complex, especially for outdoor activities. In such cases, the provision of a networking infrastructure incurs a severe environmental impact. Consider the installation of antennas to cover a wide area, and the power supply for the radios.

Circular economy

\subsection{Related works}

The research literature marginally covers the utilization of smartphones for tourist signage purposes.

An exhaustive solution is described by P. Liu in \cite{liu16a}, which provide details for an infrastructure that guides the visitors inside an urban milieu. In that case the presence of a pervasive networking facilities is a cornerstone for the whole architecture, that deeply depends on communication.

Wan \cite{wan22a} evaluates the quality signage, without making reference to a specific technology, but with many example showing phisical boards, using a formal reference, the Universal Design Principles \cite{udi97a}.

The number of research papers explodes when we extend the range to articles that investigate the impact of smartphones in tourism. The topic {\em smart tourism} topic is very popular, and covered by several review papers that provide a framework to a vast literature.

A relevant research direction covers the social aspects of related with the use of the smartphone. W. Tan \cite{tan17a} covers all aspects of a touristic experience related with the smartphone from the definition of travel destination to assistance during the visit. Much attention is dedicated to the network of connections that is established thanks to the smartphone, which again is considered in the Internet despite the title indicates a nature-based destination, where notably the Internet is not always reachable.

On the other hand, roaming in places not reached by the Internet, or {\em dead zones} using the words of Pearce and Gretzel in \cite{pea12a}, may evoke constrasting feelings, from rewarding to threatening.

More recently, the smartphone has been considered not strictly related with communication. In 2021 A.Slavec et al. investigate the use of cameras \cite{sla21a} while on travel in locations with a relevant cultural heritage in order to sustain its preservation and engage the tourist using location based games, similar to Pokemon Go or geo-caching.

In 2023 V. Rodrigues et al. publish a systematic review of papers considering the interrelationships between tourism and portable digital devices \cite{rod23a}. Although the title evokes a one-way contribution, i.e., the impact of digitalization is assumed to be positive on the quality and the sustainability of touristic offers, in the conclusions the authors reveal awareness for the necessity to address {\em "the preservation of tourism attractions/sites"} and call for for a {\em "a holistic approach ... to support a concept that still lacks conceptual and empirical clarification"}.

In that direction we meet the phenomenon of {\em overtourism}, covered by a significant literature reviewed by Dodds and Butler \cite{dod19a}, which focusses on urban tourism and its social consequences. The impact of {\em overtourism} on resorts that trade on their natural resources is investigated in the case of the Hawaii Islands \cite{lin22a} or CostaRica \cite{mat10a} stressing the imact on the social fabric.

The present paper wants to fill this gap highlighted by Rodriguez, providing a conceptual yet pragmatic approach to a well-defined aspect of tourism support taking into account sustainability by design. Once a range of relevant solutions has been identified, we proceed to the empirical part: a {\em proof of concept} implementation that verifies feasibility of a specific solution.

We stress that this article does not aim now at quantifying user satisfaction or the economic revenue associated with the specific solution. Such a target is outside the scope of our research, and indeed the figures that would represent the success of a strategy deserve further investigation. We aim at isolating an issue, propose a strategy for its solution and implement a proof of concept for it.

% Other papers examine the topic from the point of view of the user satisfaction, with an accent on the accessibility of the contents from impaired users.

% The use of smartphone is support of a tourist experience is mostly regarded from the point of view of the tourist satisfaction.

% This is the case of

% Cannot depend on communication

\subsection{A simple, low impact solution}

%Starting from previous experiences documented in the bibliography we explore a few broad categories, to find weaknesses and strengths for each of them.

We aim at designing a strategy for a sustainable and effective smartphone-based signage.

A straightforward approach concentrates on the wireless capabilities of such a device. Given the premise that the Internet is not reachable, the host might provide a local network of low power radios covering the region of interest. Small servers connected to the network might provide specific Web services. The approach requires a modest investment and a marginal environmental impact: for instance, a small device based on the ESP8266 single chip computer (SCC) has a coverage fo tens of meters, and a volume in the order of the $cm^3$. They are able to provide WiFi Access Points as well as Web contents. The ESP8266 has a capacity of 32KB, which are sufficient for an explanatory text and a low resolution image. Other SCCs, like the ESP32, are more powerful but exhibit a higher power consumption.

The dependency from power supply is a serious limitation of such kind of solution. A radio transmitter is a rather power consuming device, in the range of Watts. Even if operation is intermittent, the operation of the transmitter cannot be guaranteed for long periods and the host organization should consider the installation of a power generator, which contributes to the economical cost and environmental footprint of the device.

For this reason we do not consider a solution based on an ad-hoc network as a valid competitor against traditional board-based design for reasons related with poor sustainability.

The NFC technology is currently very diffused, and reading an NFC tag as text requires the installation of an appropriate application. Once the application is running the operation simply consists in approaching the smartphone to the tag: the tag content is transferred to the smartphone as a chink of text, and can be treated as such. The device can read it aloud to meet the user's inabilities or a translator. For both functions off-line apps exist.

Another solution in the family of passive devices is the QR-code. Such technique does not require radio communication, but uses the smartphone camera to acquire a graphical code that is translated into text. The capacity of a QR-code depends on the number of dots in the code, which in turn depends on the size of the code and the smartphone camera characteristics. We may consider that the capacity of a QR-code roughly equals that of an NFC tag with comparable size but a simpler manufacturing.

% TODO: \usepackage{graphicx} required
\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{figure/qr+NFC+coin}
	\caption[Passive devices dimensions]{An NFC coin, a QR-code card used in the project, and a 1 Euro coin by comparison on a paper with a square of 1cm}
	\label{fig:qrnfccoin}
\end{figure}

We first check the above strategies against the Universal Design Principles \cite{udi97a}:

\begin{enumerate}
	\item Equitable Use
	\item Flexibility in Use
	\item Simple and Intuitive
	\item Perceptible Information
	\item Tolerance for Error
	\item Low Physical Effort
	\item Size and Space for Approach and Use
\end{enumerate}

The labels are self-descriptive, and we address the reader to the discussion in the referenced document for a more detailed explanation.

Such minimalistic solutions (se figure \ref{fig:qrnfccoin}) compares surprisingly well when compared with other solutions that make use of the user's smartphone. There is a trade-off concerning capacity, but in many circumstances a capacity of 2-300 words is sufficient to convey the description of the site, or provide directions for the the visit. In exchange, a solution besed on NFC or QR-codes:
\begin{itemize} 
\item does not require any power supply
\item is easy to use
\item has a limited impact on the landscape
\item is not exposed to theft
\item has a negligible cost
\item is durable
\item produces a limited quantity of waste when disposed
\item is readable at a moderate distance, a feature that differentiates NFP and QR-codes
\item can be read aloud or translated
\item can be stored by the user's device.
\end{itemize}

Checking passive devices solutions against the Universal Design principles, we assess their accessibility.

\begin{itemize}
	\item {\em Equitable use} is tighly related to the smartphone technology, which is itself considered a vehicle for equitability,
	\item {\em Flexibility in use} is allowed by the computing capabilities of the device, which allows to listen to the info instead or reading, or translate the information in a different language,
	\item {\em Simple and intuitive use} holds since the operation requires an application, possibly already installed since useful in many occasions, and tag reading requires
	\item{\em Perceptible information} is a critical requirement, which contrast with the aim of keeping low the environmental intrusion. This point will be further discussed in the section devoted to the implementation
	\item{\em Tolerance for error}: there are no margins to use the device in a way that comprimizes user safety. The deliberate or accidental release of the passive device into the environment determines a minor pollution
	\item{\em Low physical effort} holds, although the user needs to carry the smartphone
	\item{Size and Space for approach and use} have to be carefully considered, since in the case of the NFC the smartphone need to be almost in touch with the passive device, while the QR-code must be in line of sight and frameable without effort
\end{itemize}

It significantly outperforms any active device, except for the capacity, and the NFC compares favorably only when capacity and size are prominent.

\section{Materials and Methods}

@@

\bibliography{biblio/paper.bib}
